---
date: "`r Sys.Date()`"
output:
    rmdformats::readthedown:
      css: readthedownstatgen.css
---

# Exercises

# **Session 1**: Operators and basic operations

> 1- In a Value for Cultivation and Use (VCU) experiment in corn, the plots consist of two four-meter rows, spaced 0.60 m apart, totaling 6 plants per row. The plant heights were respectively (in meters): 1.45; 1.73; 1.98; 2.10; 2.03; 1.65; 2.15; 1.63; 1.53; 1.98; 2.05; 1.52.

> a) Create a height vector with the above values

> b) Calculate the average height

> c) Calculate the variance of heights

```{r echo=FALSE, eval=FALSE}
ex1_vec <- c(1.45, 1.73, 1.98, 2.10, 2.03, 1.65, 2.15, 1.63, 1.53, 1.98, 2.05, 1.52)
(med_vec <- mean(ex1_vec))
(var_vec <- var(ex1_vec))
```

**Answers**:

**To view the answer, just hover over the "Spoiler alert!"**

b)[1.816667]{.spoiler}

c)[0.06540606]{.spoiler}

> 2- In the Genetics course, height and weight of 10 students were evaluated, 5 girls and 5 boys. See the table below with the respective values:

| Name       | Weight (kg) | Height (m) |
|------------|-----------|------------|
| Maria José | 72        | 1.65       |
| Mariana    | 57        | 1.68       |
| Fernanda   | 49        | 1.53       |
| Jéssica    | 62        | 1.70       |
| Carolina   | 57        | 1.52       |
| Paulo      | 90        | 1.82       |
| Marcos     | 53        | 1.63       |
| João Paulo | 76        | 1.78       |
| Geraldo    | 63        | 1.72       |
| Pedro      | 92        | 1.98       |

> a) Calculate the BMI of each student, using the formula: $BMI = \frac{weight}{height^2}$

> b) Calculate the average BMI of the class

> c) Calculate the BMI variance

> d) Calculate the BMI standard deviation

```{r echo=FALSE, eval=FALSE}
peso <- c(72,57,49,62,57,90,53,76,63,92)
altura <- c(1.65,1.68,1.53,1.7,1.52,1.82,1.63,1.78,1.72,1.98)

print("a")
(IMC <- peso/altura^2)

print("b")
(md_IMC <- mean(IMC))

print("c")
(var_IMC <- var(IMC))

print("d")
(sd_IMC <- sd(IMC))
```

**Answers**:

a)[26.44628 20.19558 20.93212 21.45329 24.67105 27.17063 19.94806 23.98687 21.29529 23.46699]{.spoiler}

b)[22.95662]{.spoiler}

c)[6.665578]{.spoiler}

d)[2.581778]{.spoiler}

> 3 - Solve the equation: $2x^2 + 8x + 6 = 0$, using the quadratic formula. Remember that:
$x = \frac{-b\pm\sqrt{\Delta}}{2a}$; $\Delta=b^2 - 4ac$

```{r echo=FALSE, eval=FALSE}
x1 <- (-8 - sqrt(8^2 - 4*2*6))/(2*2)
x2 <- (-8 + sqrt(8^2 - 4*2*6))/(2*2)

(x3 <- c(round(x1),round(x2)))
```

**Answers**:

[-3  -1]{.spoiler}


# **Session 2**: Object Classes

> 1 - Load the dataset called "mtcars" (attach(mtcars)).

> a) Check the class of the "mtcars" object

> b) Check the class of columns "hp" and "cyl"

> c) Check the class of car names (rownames)

```{r echo = FALSE, eval=FALSE}
attach(mtcars)

print("a")
(cl_mt <- class(mtcars))

print("b")
(cl_cyl_hp <- c(class(mtcars$cyl),class(mtcars$hp)))

print("c")
(cl_rnames <- class(rownames(mtcars)))
```

**Answers**:

a)["data.frame""]{.spoiler}

b)["numeric" "numeric"]{.spoiler}

c)["character"]{.spoiler}

> 2 - Using the same dataset from the previous exercise, calculate:

> a) The mean of the "hp" column

> b) Modify the class of the "hp" column to factor and calculate the mean again

> c) Compare the results obtained in letters a and b, and describe the differences between them


```{r echo = FALSE, eval=FALSE}
print("a")
(hp_mean <- mean(mtcars$hp))
print("b")
(hp_mean_fc <- mean(as.factor(mtcars$hp)))
```

**Answers**:

a)[146.6875]{.spoiler}

b)[NA e um warning]{.spoiler}

> 3 - Create a vector from a Normal distribution sample with length 3000, mean 20, and standard deviation 5. Before creating the vector, use set.seed(123456).

> a) What is the sum of observations in the vector that are greater than 30?

> b) And what is the sum of observations less than the mean? **NOTE: Calculate the mean of the set of values you just generated.**

> c) Remove the smallest observation from the vector.

```{r echo = FALSE, eval=FALSE}
set.seed(123456)
rn_val <- rnorm(n=3000, mean = 20, sd = 5)

print("a")
(rn_soma <- sum(rn_val[rn_val>30]))

print("b")
(rn_soma_sel <- sum(rn_val[rn_val<mean(rn_val)]))

print("c - apenas o 'head' ")
head((rn_rem_min <- rn_val[rn_val > min(rn_val)]))
print("c - valor mínimo:")
(min(rn_val))

```

**Answers**:

a)[2610.638]{.spoiler}

b)[23976.86]{.spoiler}

c)[only the "head": 24.16867 18.61976 18.22499 20.43744 31.26128 24.17230]{.spoiler}

c)[Minimum value: 1.170791]{.spoiler}

> 4 - In some experimental soybean areas, the presence of some nematodes was identified.

|Area | Nematode | Area yield (kg/ha)
|--------|-----------|------------|--------------
|A1      | *Meloidogyne incognita* | 2700
|A2      | *Meloidogyne incognita* | 2500
|A3      | *Heterodera glycines* | 2000
|A4      | NA | 3200
|A5      | *Pratylenchus brachyurus* | 2900
|A6      | NA | 3500
|A7      | *Meloidogyne javanica* | 2300
|A8      | NA | 3800

> a) Create a data frame with the information from the table above

> b) Create a column containing the municipality where each area is located (Piracicaba-SP, Rio Verde-GO, Londrina-PR, Barreiras-BA, Uberaba-MG, Rondonopolis-MT, Vilhena-RO, Gurupi-TO)

> c) What types of nematodes affect the evaluated areas?

> d) Which nematode had the highest impact on yield? And the lowest impact?

> e) Is the average yield of areas affected by *Meloidogyne incognita* higher than the average of areas affected by all other nematodes?


```{r echo = FALSE, eval=FALSE}

print("a")
area = c(paste0("A", rep(1:8)))
df = data.frame(c("Meloidogyne incognita", "Meloidogyne incognita", "Heterodera glycines", "NA", "Pratylenchus brachyurus", "NA", "Meloidogyne javanica", "NA"),
                c(2700, 2500, 2000, 3200, 2900, 3500, 2300, 3800), row.names = area)
colnames(df) = c("Nematode", "Produtividade (kg/ha)")
df                  

print("b")
Municipio = c("Piracicaba-SP", "Rio Verde-GO", "Londrina-PR", "Barreiras-BA", "Uberaba-MG", "Rondonopólis-MT", "Vilhena-RO", "Gurupi-TO")
df = cbind(df, Municipio)
df

print("c")
levels(df$Nematode)

print("d")
df[order(df$`Produtividade (kg/ha)`),1]

print("e") 
med_m_incog = mean(df[df$Nematode == "Meloidogyne incognita",2])
med_m_outros = mean(df[df$Nematode != "Meloidogyne incognita" & df$Nematode != "NA",2])
med_m_incog > med_m_outros
```

**Answers**:

c)["Heterodera glycines", "Meloidogyne incognita", "Meloidogyne javanica", Pratylenchus brachyurus"]{.spoiler}

d)[Highest impact: Heterodera glycines  
   Lowest impact: Pratylenchus brachyurus]{.spoiler}
   
e)[Yes. The average yield of areas with *M. incognita* (2.6 kg/ha) is higher than the average of areas affected by others (2.4 kg/ha)]{.spoiler}


# **Session 3**: Conditional and Loop Structures

1 - Time for fun! We hope you enjoyed learning about conditional structures. We didn't ask in the questionnaires about something that will be important here. Do you watch BBB (Big Brother Brazil)? Have you ever noticed that the [theme song](https://www.vagalume.com.br/bbb/vida-real-paulo-ricardo.html) of the *show* is basically a lesson in conditionals (besides being catchy, of course)? The idea is to input a verse that contains **if** and receive the next verse. If the verse doesn't exist in the song or isn't one of the conditional cases, display the message "This verse doesn't exist in the song". **Tip:** To make it easier, use all phrases in lowercase letters without accents.


```{r echo=FALSE, eval=FALSE}
if(verse=="if you could tell me"){
  "if you knew what to do"
} else if(verse == "if you knew what to do"){
  "what would you do"
} else if(verse == "if you knew who you are"){
  "how far your faith goes"
} else if(verse == "if you could choose"){
  "between good and evil"
} else if (verse == "if wanting is power"){
  "you have to go all the way"
} else if (verse == "if you want to win..."){
  " "
} else if (verse == "if I could I would take you"){
  "to where you want to go"
} else if(verse == "if you know how to play with fire"){
  "there's nothing more beautiful"
} else {
  "this verse doesn't exist in the song"
}
```

2 - Create an object "a" containing a sequence from 1 to 5. Using any loop structure, add the number 8 to each element of "a", until the sum of all elements is equal to or greater than 10000.

```{r echo=FALSE, eval=FALSE}
a = seq(1,5,1)
while(sum(a) < 10000){
  a = a + 8
}

(a)
```

**Answers**:

[2001 2002 2003 2004 2005]{.spoiler}


3 - Repita a frase "x tigre(s) triste(s) comeram x pratos de trigo", em que x seja um número inteiro da sua escolha.

```{r echo=FALSE, eval=FALSE}
# Numero de tigres/pratos
numero <- 3

# Estrutura de repetição
for(num in 1:numero){
  if(num == 1){
    print(paste(num, "tigre triste comeu", num, "prato de trigo."))
  } else {
    print(paste(num, "tigres tristes comeram", num, "pratos de trigo."))
  }
}
```

**Answers**:

Se a escolha do valor foi um:
[1 tigre triste comeu 1 prato de trigo.]{.spoiler}

Se a escolha do valor foi dois:
[1 tigre triste comeu 1 prato de trigo. 2 tigres tristes comeram 2 pratos de trigo.]{.spoiler}

4 - **Hora de descontrair!** Você conhece a música [Hey Jude](https://www.youtube.com/watch?v=7qMls5yxP1w)? Repita quantas vezes você quiser o verso:

"Na, na, na, na-na-na na

Na-na-na na, hey Jude"

```{r, echo=FALSE, eval = FALSE}
vezes <- 10

for(repete in 1:vezes){
  print("Na, na, na, na-na-na na")
  print("Na-na-na na, hey Jude")
}

```

5 - Que tal mais uma música popular? O exercício agora vai ser com essa [música](https://www.youtube.com/watch?v=C7pupFaI80A) que quase todos devem conhecer! Você irá fazer o seguinte:

* 1 - Defina o número da prole de patos
* 2 - Imprima o número da prole e a frase "patinhos foram passear"
* 3 - Adicione as seguintes linhas:
    * Além das montanhas
    * Para brincar
    * A mamãe gritou
    * *Repita o número de quacks* (**Isso difere da música, mas fica mais legal**)
* 4 - Diminua um patinho da prole, caso a prole seja menor do que um
* 5 - Imprima os versos
    * Mas só **numero da prole - 1** patinhos
    * Voltaram de lá
* 6 - Imprima linha vazia
* 7 - Retorne no passo 2 se a prole for maior que 1
* 8 - Se restar um patinho, coloque a última estrofe:

    "1 patinho foi passear

    Além das montanhas

    Para brincar

    A mamãe gritou

    Quack 

    Mas nenhum patinho

    Voltou de lá"

* 9 - Fica a seu critério colocar a última estrofe, mas de um jeito legal hein!

```{r, eval=FALSE}
prole = 3
prole_aux = prole

while(prole != 0){
  if(prole > 1){
    # Imprima o primeiro verso usando o numero de patos
     print(paste(prole, " patinhos foram passear", sep = " "))
    
    # Imprima os três versos seguintes
     print("Além das montanhas")
     print("Para brincar")
     print("A mamãe gritou")
     
    # Repita o numero de Quacks
     print(paste(rep("Quack", prole, sep=" ")))
     
    # Reduza a prole em 1
     prole = prole -1
     

    
     if(prole > 1){
       # Penultimo verso da estrofe
       print(paste("Mas só", prole, "patinhos", sep = " "))
       # Ultimo verso
       print("Voltaram de lá")
       print(" ")
     } else {
       # Penultimo verso da estrofe
       print(paste("Mas só", prole, "patinho", sep = " "))
       # Ultimo verso
       print("Voltou de lá")
       print(" ")
     }
  } else {
    # Imprima os três versos seguintes
    print("1 patinho foi passear")
    print("Além das montanhas")
    print("Para brincar")
    print("A mamãe gritou")
    print("Quack")
    print("Mas nenhum patinho")
    print("Voltou de lá")
    print(" ")
    prole = prole - 1
    print("A mamãe patinha")
    print("Foi procurar")
    print("Além das montanhas")
    print("Na beira do mar")
    print("A mamãe gritou")
    print(paste(rep("Quack", prole_aux), sep=" "))
    print(paste("E os", prole_aux, "patinhos", sep=" "))
    print("Voltaram de lá")
  }
    
}
```


6 - Utilize a expressão abaixo para gerar médias de produtividade, considerando 50 genótipos de soja, de forma aleatória:

```{r, eval = FALSE}
set.seed(123456)
produtividade = rnorm(50, mean = 3500.00, sd = 800)
```

Suponhamos que a média nacional de produtividade de soja seja de 3500.00 kg/ha. Utilizando estruturas condicionais, verifique se os genótipos apresentam valores acima ou abaixo da produtividade média nacional, retornando as frases "acima da média" ou "abaixo da média", respectivamente.

```{r echo=FALSE, eval=FALSE}
set.seed(123456)
produtividade = rnorm(50, mean = 3500.00, sd = 800)
resultado = c()
for(i in 1:length(produtividade)){
  if (produtividade[i] > 3500.00){
    resultado[i] <- ("acima da média")
  } else {resultado[i] <- ("abaixo da média")}
}

(resultado)
```

**Answers**:

[[1] "acima da média"  "abaixo da média" "abaixo da média" "acima da média"  "acima da média"  "acima da média" 
 [7] "acima da média"  "acima da média"  "acima da média"  "abaixo da média" "abaixo da média" "abaixo da média"
[13] "abaixo da média" "acima da média"  "acima da média"  "acima da média"  "abaixo da média" "acima da média" 
[19] "acima da média"  "acima da média"  "abaixo da média" "acima da média"  "acima da média"  "acima da média" 
[25] "acima da média"  "abaixo da média" "acima da média"  "acima da média"  "acima da média"  "abaixo da média"
[31] "abaixo da média" "acima da média"  "abaixo da média" "abaixo da média" "abaixo da média" "abaixo da média"
[37] "acima da média"  "acima da média"  "acima da média"  "abaixo da média" "abaixo da média" "acima da média" 
[43] "acima da média"  "abaixo da média" "acima da média"  "abaixo da média" "abaixo da média" "abaixo da média"
[49] "acima da média"  "abaixo da média"]{.spoiler}

# **Sessão 4**: Funções

Utilize o conjunto de dados disponível abaixo para gerar médias de produtividade, considerando 25 genótipos de soja avaliados em 15 ambientes, de forma aleatória:

```{r, eval=FALSE}
set.seed(123456)
dados = data.frame(Genotypes = rep(seq(1,25,1), 15), Environments = rep(x = seq(1:15), each = 25), yield = rnorm(375, 3000:3500, 800))
```

Elabore uma função para calcular as médias e o coeficiente de variação ($CV = \frac{\mu}{\sigma}*100$) em cada ambiente, e também avaliar a precisão experimental (até 20% = alta; entre 20 e 30% = média; acima de 30% = baixa). A função deve retornar os resultados contendo a identificação de cada ambiente, as médias, os CV's e as precisões.

<span style="color:red"> Pode ser que a essa altura do curso você ja tenha que ter trabalhado com diversas frustrações: o código não roda, precisou de várias tentativas até dar certo. Talvez você se sinta caindo naquele gráfico de andamento do aprendizado que mostramos na apresentação teórica. Por isso, colocamos aqui alguns vídeos motivacionais. São três palestras TED excelentes (como costuma ser os TEDs).</span>

Esses dois tem legenda em portugues. É só ajustar as configurações:

* [Teach girls bravery, not perfection | Reshma Saujani](https://www.youtube.com/watch?v=fC9da6eqaqg)
Especial para mulheres. Fala como aprender a programar nos educa a lidarmos melhor com nossas frustrações e nos desvia da cultura imposta de perfeição.

* [Mitch Resnick: Let's teach kids to code](https://www.youtube.com/watch?v=Ok6LbV6bqaE)
O palestrante é um dos desenvolvedores do programa Scratch, que consiste num jogo que ensina lógica de programação. O palestrante destaca a importância de aprender alguma linguagem de programação dos dias de hoje.

Esse só em inglês:

* [You Should Learn to Program: Christian Genco at TEDxSMU](https://www.youtube.com/watch?v=xfBWk4nw440)
Destaca os impactos de saber programar no dia-a-dia.



# **Extra**: Família dos `apply`

1 - Utilize a função abaixo para gerar médias de produtividade, considerando 25 genótipos de soja avaliados em 15 ambientes, de forma aleatória:

```{r, eval=FALSE}
set.seed(123456)
dados = data.frame(Genotypes = rep(seq(1,25,1), 15), Environments = rep(x = seq(1:15), each = 25), yield = rnorm(375, 3000:3500, 800))
```

Utilizando funções da família "apply", obtenha as médias gerais de cada genótipo e também de cada ambiente.


